EvoSwarm: A Cloud-Native Architecture for the Reproducible Execution of Multi-population Optimization Algorithms

### 2020-05-16 draft.tex
Mainly deleting addressed comments
Some rehashing of images, and more questions

% Maybe these three sections should be merged; they are all experimental set up and the two next ones start with "experimental"
% (and he next one too) - JJ
% done - Mario
% But also change title? - JJ
% Experimental design of a container-based application or something like that - JJ 
% Like it - M
\section{Experimental Design of a Container-based Application} 
\label{docker}

@@

% TODO: connection to cloud subsection  - M
% Done - M

@@

\subsubsection{Setup}
\label{setup:local-cloud}

@@

% Remember to mention cost at the end - JJ

We can define an ECS cluster directly on AWS Elastic Cloud VMs or through the AWS Fargate service.
Fargate is a serverless compute engine for containers that provides automatic provision 
and management of AWS ECS resources. With this service, calculating the cost of execution
is simplified because users only specify the computing units and memory for all the application,
instead of every component.

@@

@@

% Write some introduction to this subsubsection.
% Done - M
In this experiment, we wanted to test several deployment options for the 
cloud-native application, and have an idea of the computing capacity of each option
measured as the number of function evaluations per second.
The results are shown in Figure~\ref{fig:aws-results} and Table~\ref{tab:aws}. 
As we can see, the local environment (see section \ref{setup}) achieved a higher
ratio of evaluations per second (in bold). Still, we must consider that there is minimum
network overhead in a local workstation and a clock speed of 3.8 GHz. The first
option (labeled Single) is straightforward to deploy, needing only a single
{\tt docker-compose} file, and one command to execute. But with the limit of 4 CPUs and
ten containers, it is the slowest option. The second option (labeled Multi), is
faster, and not too far from the local execution, especially when solving the
$f_4$ in 20 dimensions (underlined in table). With the capacity for increasing the number of worker
containers, it is reasonable to think that we can achieve a higher number of
evaluations with an appropriate configuration. Nevertheless, the goal of this
experiment was not to increase the number of evaluations but to test the
flexibility of a cloud-native application. The total cost of testing, configuration, 
and execution of the experiments in ECS was less than \$1.00 US Dollar.

% Mention costs - JJ
% Done  - M

@@

% Emphasize these results for the purpose of the paper - JJ
% done - Mario

@@

% You should clarify that, before, you were explaining how to set up
% an experiment. In this section, you are explaining the experiments
% that you are doing in this paper. But you should relate both
% sections to the general objectives of the paper. In the prior
% section, you want to prove that EvoSwarm can be used by the public
% to create reproductive research (and maybe you should add a small
% conclusion sub or sub-sub section), in the next you explain the
% experiment we are doing for this paper. It would be interesting to
% make references from the next section to the previous one,
% indicating how the general setup of EvoSwarm helps _us_ to create
% this experiment - JJ  
% Connected below - M

@@

% Also, point forward to the section (or subsections) that will be devoted to proving each point.
% BTW, they should probably be subsections.
% Done - M

@@

Next, we show the parameters for each algorithm, with Table \ref{tab:GAparams} for
the GA and Table \ref{tab:PSOparams} for PSO. We obtained these parameters
following the same method as in a previous work \cite{garcia2017benchmarking}.
To obtain the parameters, we tested first on the Rastrigin separable function
with five dimensions. After about fifteen experiments, the most challenging
targets were achieved for this particular function. We tested again with
functions one to three, and after obtaining favorable results, the PSO and GA
algorithm parameters were set. In the GA, we randomly set the mutation and
crossover probabilities to have more heterogeneous workers; % highlight this and add citation - JJ 
% done below - M
in these parameters, the range of values is specified. We did not change these parameters during the 
experiments, and only the population size and number of generations were
provided as parameters. This strategy has been applied successfully for increasing 
the performance of multi-population algorithms without requiring the setting 
of initial parameters \cite{garcia2014randomized}.

@@

% Maybe subsection? Also say the question you're answering - JJ
% Done, I added the research question at the beginning of the setup question - M
\subsection{Measuring Scalability}
\label{sec:exp1}
% Experiment 1

@@

@@

%
%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{evalspersecond}
  \caption{Boxplot of the number of evaluations per second, for 30 instances of function $f_4$, with different number of workers and populations sizes }
  \label{fig:spworker:evalspersecond}
\end{figure}

  \begin{tabular}{c|r|r|r|r|r|r|}
  \cline{2-7}
  \multicolumn{1}{l|}{}           & \multicolumn{3}{c|}{5 Populations}                                          & \multicolumn{3}{c|}{10 Populations}                                         \\ \hline
  \multicolumn{1}{|l|}{Dimension} & \multicolumn{1}{c|}{2W} & \multicolumn{1}{c|}{4W} & \multicolumn{1}{c|}{8W} & \multicolumn{1}{c|}{2W} & \multicolumn{1}{c|}{4W} & \multicolumn{1}{c|}{8W} \\ \hline
  \multicolumn{1}{|c|}{2}         & \textbf{1.94}           & 3.12                    & 3.86                    & 1.84                    & \textbf{3.38}                    & \textbf{5.98}                    \\ \hline
  \multicolumn{1}{|c|}{3}         & \textbf{1.93}           & 2.77                    & 3.15                    & 1.90                    & \textbf{3.26}                   & \textbf{5.96}                    \\ \hline
  \multicolumn{1}{|c|}{5}         & 1.91                    & 2.90                    & 3.09                    & 1.91                    & \textbf{3.46}                    & \textbf{6.14}                    \\ \hline
  \multicolumn{1}{|c|}{10}        & \textbf{1.92}           & 3.03                    & 3.30                    & 1.90                    & \textbf{3.61}                   & \textbf{6.41}                    \\ \hline
  \multicolumn{1}{|c|}{20}        & \textbf{1.91}           & 3.14                    & 3.27                    & 1.89                    & \textbf{3.73}                   & \textbf{6.71}                    \\ \hline
  \multicolumn{1}{|c|}{40}        & \textbf{2.00}           & 2.92                    & 3.12                    & 1.95                    & \textbf{3.80}                     & \textbf{7.52}                    \\ \hline
  \end{tabular}
\end{table}
%

\begin{table}[h!tbp]
  \small
  \caption{Speedup achieved with incrementing the number of workers for
   each dimension showing the p-rate for the Wilcoxon test}
  % mark significant differences in some way - JJ
  \label{tab:speedup:test}
  \vspace{0.25cm}
  \centering
  \begin{tabular}{lllllll}
            &               &               & \multicolumn{2}{l}{5 Populations} & \multicolumn{2}{l}{10 Populations} \\
  Dimension & \multicolumn{2}{l}{Increment} & Speedup         & p-value         & Speedup          & p-value         \\
  2         & 1w            & 2w            & {\ul 1.94}            & 6E-16           & 1.84             & 2E-16           \\
            & 2w            & 4w            & 1.61            & 2E-15           & 1.83             & 1E-15           \\
            & 4w            & 8w            & 1.24            & 1E-06           & 1.77             & 2E-16           \\
  3         & 1w            & 2w            & {\ul 1.93}            & 2E-16           & 1.90             & 2E-16           \\
            & 2w            & 4w            & 1.44            & 8E-15           & 1.71             & 2E-16           \\
            & 4w            & 8w            & 1.14            & 3E-05           & 1.83             & 2E-16           \\
  5         & 1w            & 2w            & 1.91            & 9E-16           & 1.91             & 2E-16           \\
            & 2w            & 4w            & 1.52            & 3E-11           & 1.82             & 2E-16           \\
            & 4w            & 8w            & 1.06            & 5E-03           & 1.77             & 2E-16           \\
  10        & 1w            & 2w            & 1.92            & 3E-16           & 1.90             & 2E-16           \\
            & 2w            & 4w            & 1.58            & 3E-16           & 1.89             & 2E-16           \\
            & 4w            & 8w            & {\ul 1.09}      & 1E-03           & 1.78             & 2E-16           \\
  20        & 1w            & 2w            & 1.91            & 2E-16           & 1.89             & 2E-16           \\
            & 2w            & 4w            & 1.64            & 2E-15           & 1.97             & 2E-16           \\
            & 4w            & 8w            & {\ul 1.04 }     & 1E-02           & 1.80             & 2E-16           \\
  40        & 1w            & 2w            & \textbf{2.00}   & 2E-16           & 1.95             & 2E-16           \\
            & 2w            & 4w            & 1.46            & 1E-09           & 1.95             & 2E-16           \\
            & 4w            & 8w            & {\ul 1.07}            & 3E-02           & \textbf{1.98}    & 2E-16          
  \end{tabular}
  \end{table}

@@

However, when we compare the two
options, we find some subtle differences: When using one or two
workers, using 5 populations takes less time in most cases. 
In Table~\ref{tab:time} results with p-value of the two-tailed Wilcoxon
signed-ranks test was grater than the level of significance of 0.05 
(i.e., accepting the null hypothesis of inequality) are marked in boldface.
A reason for this could be that the migration of individuals takes more time to
reach the worker. For instance, when a single worker has 10 populations, only
after processing every one of them, it will start receiving mixed populations.  Although, in
all cases, and even a single worker, populations are heterogeneous, this
advantage takes more time to propagate. With four workers, almost the same
results can be achieved regardless of our selection between five or ten
populations. Also, there is not much difference between 4 and 8 workers, when
using five populations, which might indicate that some workers are
idling while populations are being processed by other workers. In
general, we can observe that keeping a ratio close to 1:1, but a bit
better than that, between the number of workers and populations, better exploits the computing
resources available. This also decreases the relation between the
overhead needed to migrate and process incoming and outgoing
population and the total computing time: more populations will need
more time to do that.

\subsubsection{Speedup}
\label{sec:speedup}

% TBD? - JJ

To measure the speedup obtained by the addition of workers, we will use the
number of evaluations per second, instead of the time required to finish an
instance of a problem. The reason for this is that in some cases reaching the
desired target can require less work. We are going to measure the total increase
of work done in a second by adding more workers. The baseline for the comparison
will be one worker, and the minimum amount of work available will be five
populations. Five populations are selected because the controller needs at least
three populations to migrate individuals between them, and we add two more to
keep even a single worker busy.Figure \ref{fig:spworker:evalspersecond} 
shows the number of function evaluation per second, achieved by a certain number of
workers, for every problem dimension, for five and ten dimensions of $f_4$.

As expected, within the same number of populations, there is a linear speedup
when we add more workers. The speedup is marginal in the case of five
populations when increasing the number of workers above 4. The reason for this
is that at any given time, there are at least two populations in the message
queue, leaving four workers waiting. We can also notice that the spread in the
amount of work is less when using up to four workers.

Table~\ref{tab:speedup-table} shows the speedup obtained against the baseline of
one worker. When using eight workers, we achieved the maximum speed up for five
and ten populations, but interestingly, it was in dimension 40 and dimension 2 
(these values are shown in bold in the table).

Table~\ref{tab:speedup:test} shows the speedup obtained taking as the baseline
the previous configuration, for example, when increasing from four to eight
workers, and the p-rate for the left-tailed Wilcoxon test. There was an
increment in all cases, but again, for smaller dimensions, five populations with
fewer workers is a better option (see values underlined in the table) with the best  
for 40 dimensions (in bold). The worst speed was going from four to eight workers 
at higher dimensions (underlined in the table). For ten populations, overall 
has the linear speedup but the best is from four to eight workers at 40 dimensions
(shown in bold).

@@

>>>>

\section{Experimental Design of a Container-based Application} 
\label{docker}

@@

@@

\subsubsection{Setup} % Even if it's a sub-sub section, flesh this out ... - JJ
\label{setup:local-cloud}

@@

% Remember to mention cost at the end - JJ
% Ping - JJ
We can define an ECS cluster directly on AWS Elastic Cloud VMs or through the AWS Fargate service.
Fargate is a serverless compute engine for containers that provides automatic provision 
and management of AWS ECS resources. With this service, calculating the cost of execution
is simplified because users only specify the computing units and memory for all the application,
instead of every component.

@@

% Mention cost here - JJ

@@

In this experiment, we wanted to test several deployment options for the 
cloud-native application, and have an idea of the computing capacity of each option
measured as the number of function evaluations per second.
The results are shown in Figure~\ref{fig:aws-results} and Table~\ref{tab:aws}. 
As we can see, the local environment (see section \ref{setup}) achieved a higher
ratio of evaluations per second (in bold). Still, we must consider that there is minimum
network overhead in a local workstation and a clock speed of 3.8 GHz. The first
option (labeled Single) is straightforward to deploy, needing only a single
{\tt docker-compose} file, and one command to execute. But with the limit of 4 CPUs and
ten containers, it is the slowest option. The second option (labeled Multi), is
faster, and not too far from the local execution, especially when solving the
$f_4$ in 20 dimensions (underlined in table). With the capacity for increasing the number of worker
containers, it is reasonable to think that we can achieve a higher number of
evaluations with an appropriate configuration. Nevertheless, the goal of this
experiment was not to increase the number of evaluations but to test the
flexibility of a cloud-native application. The total cost of testing, configuration, 
and execution of the experiments in ECS was less than \$1.00 US Dollar.

@@

@@

@@

@@

Next, we show the parameters for each algorithm, with Table \ref{tab:GAparams} for
the GA and Table \ref{tab:PSOparams} for PSO. We obtained these parameters
following the same method as in a previous work \cite{garcia2017benchmarking}.
To obtain the parameters, we tested first on the Rastrigin separable function
with five dimensions. After about fifteen experiments, the most challenging
targets were achieved for this particular function. We tested again with
functions one to three, and after obtaining favorable results, the PSO and GA
algorithm parameters were set. In the GA, we randomly set the mutation and
crossover probabilities to have more heterogeneous workers; 
in these parameters, the range of values is specified. We did not change these parameters during the 
experiments, and only the population size and number of generations were
provided as parameters. This strategy has been applied successfully for increasing 
the performance of multi-population algorithms without requiring the setting 
of initial parameters \cite{garcia2014randomized}.

@@

\subsection{Measuring Scalability}
\label{sec:exp1}
% Experiment 1

@@

\subsubsection{Setup} % Again, this in the index with no context should say something. Please flesh it out.

@@

@@

However, when we compare the two
options, we find some subtle differences: When using one or two
workers, using 5 populations takes less time in most cases. 
In Table~\ref{tab:time} results with p-value of the two-tailed Wilcoxon
signed-ranks test was grater than the level of significance of 0.05 
(i.e., accepting the null hypothesis of inequality) are marked in boldface.
A reason for this could be that the migration of individuals takes more time to
reach the worker. % how come? - JJ
For instance, when a single worker has 10 populations, only
after processing every one of them, it will start receiving mixed populations.  Although, in
all cases, and even with a single worker, populations are heterogeneous, this
advantage takes more time to propagate. % which advantage? - JJ
With four workers, almost the same
results can be achieved regardless of our selection between five or ten
populations. Also, there is not much difference between 4 and 8 workers, when
using five populations, which might indicate that some workers are
idling while populations are being processed by other workers. In
general, we can observe that keeping a ratio between the number of workers and populations a bit over 1:1  exploits better the computing
resources available. This also decreases the relation between the
overhead needed to migrate and process incoming and outgoing
population and the total computing time: more populations will need
more time to do that.

%
%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{evalspersecond}
  \caption{Boxplot of the number of evaluations per second, for 30 instances of function $f_4$, with different number of workers and populations sizes }
  \label{fig:spworker:evalspersecond}
\end{figure}
%
\subsubsection{Speedup}
\label{sec:speedup}

% Introduction saying why we are measuring this, how it's related to the objectives of the paper - JJ
To measure the speedup obtained by the addition of workers, we will use the
number of evaluations per second, instead of the time required to finish an
instance of a problem. The reason for this is that in some cases reaching the
desired target can require less work. We are going to measure the total increase
of work done in a second by adding more workers. The baseline for the comparison
will be the algorithm with a single worker, and the minimum amount of work available will be five
populations. This number is selected because the controller needs at least
three populations to migrate individuals between them, and we add two more to
keep even a single worker busy. Figure \ref{fig:spworker:evalspersecond} 
shows the number of function evaluation per second, achieved by a certain number of
workers, for every problem dimension, for five and ten dimensions of $f_4$.  % Why did you choose f4? - JJ

  \begin{tabular}{c|r|r|r|r|r|r|}
  \cline{2-7}
  \multicolumn{1}{l|}{}           & \multicolumn{3}{c|}{5 Populations}                                          & \multicolumn{3}{c|}{10 Populations}                                         \\ \hline
  \multicolumn{1}{|l|}{Dimension} & \multicolumn{1}{c|}{2W} & \multicolumn{1}{c|}{4W} & \multicolumn{1}{c|}{8W} & \multicolumn{1}{c|}{2W} & \multicolumn{1}{c|}{4W} & \multicolumn{1}{c|}{8W} \\ \hline
  \multicolumn{1}{|c|}{2}         & \textbf{1.94}           & 3.12                    & 3.86                    & 1.84                    & \textbf{3.38}                    & \textbf{5.98}                    \\ \hline
  \multicolumn{1}{|c|}{3}         & \textbf{1.93}           & 2.77                    & 3.15                    & 1.90                    & \textbf{3.26}                   & \textbf{5.96}                    \\ \hline
  \multicolumn{1}{|c|}{5}         & 1.91                    & 2.90                    & 3.09                    & 1.91                    & \textbf{3.46}                    & \textbf{6.14}                    \\ \hline
  \multicolumn{1}{|c|}{10}        & \textbf{1.92}           & 3.03                    & 3.30                    & 1.90                    & \textbf{3.61}                   & \textbf{6.41}                    \\ \hline
  \multicolumn{1}{|c|}{20}        & \textbf{1.91}           & 3.14                    & 3.27                    & 1.89                    & \textbf{3.73}                   & \textbf{6.71}                    \\ \hline
  \multicolumn{1}{|c|}{40}        & \textbf{2.00}           & 2.92                    & 3.12                    & 1.95                    & \textbf{3.80}                     & \textbf{7.52}                    \\ \hline
  \end{tabular}
\end{table}
%
As expected, within the same number of populations, there is a linear speedup
when we add more workers. The speedup is marginal in the case of five
populations when increasing the number of workers above 4. The reason for this
is that at any given time, there are at least two populations in the message
queue, leaving four workers waiting. We can also notice that the spread in the
amount of work is less when using up to four workers.

\begin{table}[h!tbp]
  \small
  \caption{Speedup achieved with incrementing the number of workers for
    each dimension showing the p-rate for the Wilcoxon test.}
  % Say what highlights mean - JJ
  \label{tab:speedup:test}
  \vspace{0.25cm}
  \centering
  \begin{tabular}{lllllll}
            &               &               & \multicolumn{2}{l}{5 Populations} & \multicolumn{2}{l}{10 Populations} \\
  Dimension & \multicolumn{2}{l}{Increment} & Speedup         & p-value         & Speedup          & p-value         \\
  2         & 1w            & 2w            & {\ul 1.94}            & 6E-16           & 1.84             & 2E-16           \\
            & 2w            & 4w            & 1.61            & 2E-15           & 1.83             & 1E-15           \\
            & 4w            & 8w            & 1.24            & 1E-06           & 1.77             & 2E-16           \\
  3         & 1w            & 2w            & {\ul 1.93}            & 2E-16           & 1.90             & 2E-16           \\
            & 2w            & 4w            & 1.44            & 8E-15           & 1.71             & 2E-16           \\
            & 4w            & 8w            & 1.14            & 3E-05           & 1.83             & 2E-16           \\
  5         & 1w            & 2w            & 1.91            & 9E-16           & 1.91             & 2E-16           \\
            & 2w            & 4w            & 1.52            & 3E-11           & 1.82             & 2E-16           \\
            & 4w            & 8w            & 1.06            & 5E-03           & 1.77             & 2E-16           \\
  10        & 1w            & 2w            & 1.92            & 3E-16           & 1.90             & 2E-16           \\
            & 2w            & 4w            & 1.58            & 3E-16           & 1.89             & 2E-16           \\
            & 4w            & 8w            & {\ul 1.09}      & 1E-03           & 1.78             & 2E-16           \\
  20        & 1w            & 2w            & 1.91            & 2E-16           & 1.89             & 2E-16           \\
            & 2w            & 4w            & 1.64            & 2E-15           & 1.97             & 2E-16           \\
            & 4w            & 8w            & {\ul 1.04 }     & 1E-02           & 1.80             & 2E-16           \\
  40        & 1w            & 2w            & \textbf{2.00}   & 2E-16           & 1.95             & 2E-16           \\
            & 2w            & 4w            & 1.46            & 1E-09           & 1.95             & 2E-16           \\
            & 4w            & 8w            & {\ul 1.07}            & 3E-02           & \textbf{1.98}    & 2E-16          
  \end{tabular}
  \end{table}
%
Table~\ref{tab:speedup-table} shows the speedup obtained against the baseline of
one worker. When using eight workers, we achieved the maximum speed up for five
and ten populations, but interestingly, it was in dimension 40 and dimension 2 
(these values are shown in bold in the table).

Table~\ref{tab:speedup:test} shows the speedup obtained taking as the baseline
the previous configuration, for example, when increasing from four to eight
workers, and the p-rate for the left-tailed Wilcoxon test. There was an
increment in all cases, but again, for smaller dimensions, five populations with
fewer workers is a better option (see values underlined in the table) with the best speedup, achieved for 40 dimensions, shown in bold. The worst speed was going from four to eight workers 
at higher dimensions (underlined in the table). For ten populations, overall 
has the linear speedup % don't understand this - JJ
but the best is from four to eight workers at 40 dimensions
(shown in bold).

@@

% Write an introduction - JJ
